---
title: "Democratic Backsliding in Latin America"
subtitle: "How has electoral democracy changed globally and across selected Latin American countries between 2010 and 2023?"
author: "Ally Denton"
institute: "American University"
format: 
  beamer:
    theme: Singapore
    colortheme: seagull
    aspectratio: 43
knitr: 
  opts_chunk: 
    echo: false
    eval: true
    prompt: false
    tidy: false
    message: FALSE
    warning: FALSE
    fig.align: "center"
fontsize: 10pt
---


```{r}

library(tidyverse)
library(haven)
library(knitr)
library(ggrepel)
library(maps)

```

## Overview 

Democratic backsliding has become a central concern in international affairs. As such, this project examines global change in electoral democracy from 2010–2023, with a focus on variation across four Latin American countries.

**Central Findings**

-  Democratic trajectories vary across countries due to differences in domestic leadership and external political and economic conditions.
-  Global electoral democracy has declined modestly over the 2010–2023 period.
-  Latin American cases show divergent trajectories, with both decline and improvement depending on the country.

## Data

Dataset: Varieties of Democracy 

-  The dataset is structured as country-year panel data, with one observation per country-year.
-  For this project, I am specifically looking at the span of 2010-2023 to try and optimize the data available. 
- The Data will cover all countries that have data over this time period and also zoom in on four Latin American cases: Venezuela, Brazil, Ecuador, El Salvador.
    - These cases were chosen to reflect variation in regime type, economic performance, and political stability.
-  The core variable for this project is: 
    - the v2x_polyarchy variable (the Electoral Democracy Index) 
        - The variable measures the extent to which the ideal of electoral democracy in its fullest sense is achieved. 
        - The variable itself uses a 0-1 scale with 0 indicating low electoral democracy and 1 indicating high electoral democracy. 

## Trends in Overall Democracy Over Time
```{r}
#| label: tbl-avgall
#| fig-width: 8
#| fig-asp: 0.67
#| tbl-cap: "Average Electoral Democracy over Time"

library(knitr)

vdem <- read_csv('vdem.csv')

vdem %>%
  group_by(year) %>%
  filter(year == 2010 | year == 2023) %>%
  summarize(
    mean = mean(v2x_polyarchy, na.rm = T)
  ) %>%
  kable(digits = 2L)

```

- Based on @tbl-avgall, electoral democracy in the world on average has seen a decrease of .031 points on the electoral democracy index in the past decade. 

## Electoral Democracy Throughout the World
```{r}
#| label: fig-map
#| fig-width: 8
#| fig-asp: 0.4
#| fig-cap: "Change in Electoral Democracy across the World (2010–2023)"

# transforming the data

vdem_change <- vdem %>%
  filter(year %in% c(2010, 2023)) %>%
  select(country_name, year, v2x_polyarchy) %>%
  pivot_wider(
    names_from = year, 
    values_from = v2x_polyarchy) %>%
  mutate(change = `2023` - `2010`,
         country_name = case_when(
           country_name == 'United States of America' ~ 'USA',
           TRUE ~ country_name))

dmap <- vdem_change %>%
  left_join(
    map_data('world'),
    by = c('country_name' = 'region')
  )


# visualization

gradient.map <- dmap %>%
ggplot(aes(x = long, y = lat, group = group)) +
  geom_polygon(aes(fill = change), color = "white", size = 0.1) +
  scale_fill_gradient2(
    low = 'darkred',
    mid = 'lightgray',
    high = 'navy',
    midpoint = 0) +
  coord_fixed(1.3) +
  labs(
    fill = 'Change',
    subtitle = 'Red = Backsliding | Blue = Improvement',
    caption = 'Source: V-Dem') +
   theme_void() +
   theme(
     plot.title = element_text(face = 'bold')
   )

gradient.map

```
- The global pattern indicates a small decline in electoral democracy over the period, though outcomes differ significantly across countries and regions. 

## Zooming in on the Latin American Cases
```{r}
#| label: tbl-avgLA
#| tbl-width: 8
#| tbl-asp: 0.67
#| tbl-cap: "Average Electoral Democracy over Time"

latin_cases <- c('Venezuela', 'Brazil', 'Ecuador', 'El Salvador')

vdem %>%
  group_by(year, country_name) %>%
  select(country_name, year, v2x_polyarchy) %>%
  filter(year == 2010 | year == 2023) %>%
  filter(country_name %in% latin_cases) %>%
  summarize(
    mean = mean(v2x_polyarchy, na.rm = T)
  ) %>%
  pivot_wider(
    names_from = year,
    values_from = mean
  ) %>%
  kable(digits = 2L)

```
- Looking at @tbl-avgLA, the average electoral democracy does decrease for the majority of these cases. 
- Ecuador is the main exception, showing a net increase over the period.
- Brazil remains relatively high on the electoral democracy scale despite mild fluctuation.

## The Democracy Trends across Latin America over Time
```{r}
#| label: fig-LA
#| fig-width: 10
#| fig-asp: 0.4
#| fig-cap: "Democratic Trajectories in Latin America from 2010 to 2023"

demtraj <- vdem %>%
  filter(country_name %in% latin_cases,
         year >= 2010) %>%
  ggplot(aes(x = year, y = v2x_polyarchy, color = country_name, linetype = country_name)) +
  geom_line(size = 1) +
  scale_color_manual(values = c('springgreen4', 'gold', 'blue3', 'red2')) +
  scale_linetype_manual(values = c(
    'Brazil' = 'longdash',
    'Ecuador' = 'solid',
    'El Salvador' = 'dashed',
    'Venezuela' = 'dotdash')) +
  scale_x_continuous (
    limits = c(2010, 2025), 
    breaks = seq(2010, 2025, 5),
    expand = expansion(mult = c(0, .1))
                       ) +
  scale_y_continuous(
    limits = c(0,1),
    breaks = seq(0, 1, .2),
    expand = expansion(mult = c(0, .1))
  ) +
  labs(
    y = 'Electoral Democracy Index',
    x = NULL,
    color = 'Country',
    linetype = 'Country'
  ) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = 'gray90'),
        plot.title = element_text(face = 'bold'),
        panel.background = element_rect(fill = 'white'))

demtraj

```
- Looking at @fig-LA, the Latin American cases do not follow a uniform regional pattern. 
    - Venezuela and El Salvador exhibit sustained declines in electoral democracy, while Brazil and Ecuador show more variation, including periods of recovery or relative improvement.

## Conclusion

- Overall, the evidence suggests a small decline in electoral democracy globally between 2010 and 2023, though the magnitude of change is limited.

- At the regional level, Latin America does not follow a single trajectory.
    - Venezuela and El Salvador show sustained declines.
    - Brazil and Ecuador display more mixed trajectories, including periods of improvement.

**Overall, these findings highlight that democratic change is not uniform across countries or regions. Rather, electoral democracy evolves through distinct national trajectories, even within the same geographic region.**

